% $Id: report-2009-03.tex 4103 2009-03-18 11:58:46Z joostvb $
% $URL: svn+ssh://nagy/data/vc/svn/trunk/doc/nlnet-mailman-pgp-smime/report-2009-03.tex $
%
% Copyright © 2009, 2010 Joost van Baal <info@ad1810.com>
%

\documentclass[a4]{article}
\usepackage{a4wide}
\usepackage{url}

% \title{First Secure List Server bi-monthly project report\\September 1, 2008}
\title{Sixth Secure List Server project report}
% Jan 1, 2009

\author{Joost van Baal \url{<joostvb@ad1810.com>}}

\begin{document}

\maketitle

\setlength{\parindent}{0pt}
\setlength{\parskip}{1.5ex}

\section*{Introduction}

As agreed upon in the June 13, 2008 document ``Memorandum of Understanding
Secure List Server Project'', this report documents work done by the author for
the Secure List Server project
(\url{http://non-gnu.uvt.nl/mailman-pgp-smime/}), as funded by the NLnet
Foundation.  It also lists the current plans for the project.

This document is a follow-up to the Fifth Secure List Server
project report, as sent to Valer Mischenko on 24 February 2010.

%downloads, list traffic, commits

\section{Completed tasks}

Here's a condensed overview of the progress made thus far.

\begin{tabular}{lll}


 Task                                Planned   Delivered
 (Start project)                     08-07-01  08-06-24
 (Milestone 1)                       08-08-15  08-08-09
 (Milestone 2)                       08-12-15  09-01-06
 (Milestone 3)                       09-03-01  09-09-09


 Task                            & Planned  & Delivered   \\ \cline{1-3}
 (Start project)                 & 08-07-01 & 08-06-24    \\
 (Milestone 1)                   & 08-08-15 & 08-08-09    \\
 (Milestone 2)                   & 08-12-15 & 09-01-06    \\
 Publish third project report    & 09-01-01 & 09-01-26    \\
 Write and publish documentation & 09-01-15 & 09-01-12    \\
 Create a package of SLS         & 09-03-01 & in progress \\
 Publish fourth project report   & 09-03-01 & 09-03-18    \\
 Disseminate results             & 09-03-01 & 09-02-08    \\
 (Milestone 3)                   & 09-03-01 & in progress \\
\end{tabular}

See the previous reports for details on the tasks completed for
Milestones 1, 2 and 3.



Some extra time was spent on keeping our code in sync with upstream by merging
it.  Feedback has been given to users on the GPG/SMIME Mailman development
list.


\section{Planned tasks}

Current plans are:

merge datenritter patches!

Date: Tue, 06 Oct 2009 13:34:37 +0200
From: datenritter <lists datenritter.de>
To: joostvb-mailman-developers
Subject: Re: [Ssls-dev] missed a patch in last release
Message-ID: <4ACB2B4D.5090604@datenritter.de>



                                     Planned
 Act upon auditors final report      10-03-18
 Try get SLS shipped w/ distros      10-03-29
 Fifth and final project report      10-03-29
 (Milestone 4)                       10-03-29



\begin{tabular}{lll}
 Task                            & Planned  \\ \cline{1-2}
 Create a package of SLS         & 09-03-01 \\
 (Milestone 3)                   & 09-03-01 \\
\end{tabular}

The final audit report mentioned the lax checking of multipart/signed messages.
This has been fixed.  Furthermore, it suggested to reduce the number of
configuration options available to the list administrator.  I'll work on this.

The current plan is:
                                     Planned
 Act upon auditors final report      10-03-18
 Try get SLS shipped w/ distros      10-03-29
 Fifth and final project report      10-03-29
 (Milestone 4)                       10-03-29

Originally, completion of Milestone 4 was planned for April 15, 2009.

Next to the listed tasks, I'll port the patch to upstream mailman 2.1.13
(released 09-12-22).  Furthermore, I'll update the RPM and .deb packages to
this new upstream release.

I'll contact Mailman developer Barry Warsaw and ask him to perform his review
he has offered via the Mailman Developers list.

--------------

Message-ID: <PC1952010030220411002187fb4e6a4@msapiro>
Date: Tue, 2 Mar 2010 20:41:10 -0800
From: Mark Sapiro
To: mailman-developers@python.org
In-Reply-To: <4f6fd0b11003022002u34652a5bnb86decfa0102568b@mail.gmail.com>
Subject: Re: [Mailman-Developers] Implementing a new feature, first steps

>I am interested in implementing a new feature in Mailman

>Of course, any code I write (and which you find useful) will go right back
>upstream!

Thank you. However, it won't go in 2.1.x, because that's bug fix only
and it probably won't be applicable, at least directly, to MM 3.

--------------

I'll contact responsible parties and try to get SLS shipped with Debian, Ubuntu
and an RPM-based distribution (like Fedora) as well as the Sabayon and
Smallsister projects.


\end{document}

