% $Id: report-2008-11.tex 3767 2008-11-01 18:05:18Z joostvb $
% $URL: svn+ssh://nagy/data/vc/svn/trunk/doc/nlnet-mailman-pgp-smime/report-2008-11.tex $
%
% Copyright © 2008 Joost van Baal <info.ad1810.com>
%

\documentclass[a4]{article}
\usepackage{a4wide}
\usepackage{url}

% \title{First Secure List Server bi-monthly project report\\September 1, 2008}
\title{Second Secure List Server bi-monthly project report}
% November 1, 2008

\author{Joost van Baal \url{<joostvb@ad1810.com>}}

\begin{document}

\maketitle

\setlength{\parindent}{0pt}
\setlength{\parskip}{1.5ex}

\section*{Introduction}

As agreed upon in the June 13, 2008 document ``Memorandum of Understanding
Secure List Server Project'', this report documents work done by the author for
the Secure List Server project
(\url{http://non-gnu.uvt.nl/mailman-pgp-smime/}), as funded by the NLnet
Foundation.  It also lists the current plans for the project.

This document is a follow-up to the First Secure List Server bi-monthly project
report, as sent to Valer Mischenko on Aug 31, 2008.

\section{Completed tasks}

Here's a condensed overview of the progress made thus far.

\begin{tabular}{lll}
 Task                            & Planned  & Delivered   \\ \cline{1-3}
 (Start project)                 & 08-07-01 & 08-06-24    \\
 (Milestone 1)                   & 08-08-15 & 08-08-09    \\
 Publish first project report    & 08-09-01 & 08-08-31    \\
 Bug: Implement test suite       & 08-10-15 & 08-10-30    \\
 Publish second project report   & 08-11-01 & 08-11-01    \\
\end{tabular}

See the first report for details on the tasks completed for Milestone 1.

The test suite consists of two parts: first a simple script
pgp-smime/pgp-smime-testsuite.sh, which was finished with bzr commit revno: 460
on branch nick 2.1-pgp-smime.  See the Mailman pgp-smime bzr branch in
Launchpad at \url{https://code.launchpad.net/~joostvb/mailman/2.1-pgp-smime}.
The second part of the suite is a running SLS installation on the author's
workstation, hosting several test lists with additional PGP and S/MIME key
pairs.  Work on documenting the setup of such a system is currently in progress
(to be completed with other documentation on 09-01-15).

Some extra time was spent on releasing version 2008-09-25 of the patch,
available from
\url{http://non-gnu.uvt.nl/pub/mailman/mailman-2.1.11-pgp-smime_2008-09-25.patch.gz},
announced in \url{Message-ID: <20080925084757.GK14261@bruhat.mdcc.cx>} on the
GPG/SMIME mailman devel list at Thu, 25 Sep 2008.  Work on shipping a new
release is currently in progress, it'll likely get shipped this month.

Contacting Sabayon's Fabio Erculiani took some time (15 Oct 2008).  There has
also been contact with Brenno de Winter (the Smallsister project, 13 Oct 2008).

A minor status update on SLS has been posted to the Mailman Developers (28 Sep
2008, \url{Message-ID: <20080928110912.GT11325@bruhat.mdcc.cx>}, archived at
\url{http://www.mail-archive.com/mailman-developers%40python.org/msg11143.html}.

Finally, I've acted upon helpful feedback from Lars Kruse about the SLS
website.

\section{Planned tasks}

\begin{tabular}{lll}
 Task                            & Planned  \\ \cline{1-2}
 Bug: Enforce confidentiality    & 08-11-15 \\
 Bug: Better user interface      & 08-12-15 \\
 (Milestone 2)                   & 08-12-15 \\
 Publish third project report    & 09-01-01 \\
 Write and publish documentation & 09-01-15 \\
 Create a package of SLS         & 09-03-01 \\
 Publish fourth project report   & 09-03-01 \\
 Disseminate results             & 09-03-01 \\
 Try get SLS shipped w/ distros  & 09-03-01 \\
 (Milestone 3)                   & 09-03-01 \\
 Act upon auditors final report  & 09-04-01 \\
 (Milestone 4)                   & 09-04-01 \\
 Fifth and final project report  & 09-04-01 \\
\end{tabular}

The 2 open Bug-tasks are critical open bugs as found by Security Auditor
Guus Sliepen and published in ``Security Audit of the Secure List Server, Part
I'', August 1, 2008.  While fixing these bugs, the patch will be kept as
non-intrusive and minimal as possible.

Once these bugs are fixed, I'll contact Mailman developer Barry Warsaw and ask
him to perform his review he has offered via the Mailman Developers list.

The documentation will be written for users, for list admins, for site admins,
as well as for developers.

Both a Debian and an RPM package for SLS will get build and published.

The dissemination of results will be done by announcing the achievements on
relevant mailing lists, and by giving a presentation at e.g. the FOSDEM
conference (will probably happen february 2009 in Brussels), and/or
a meeting of CCC Ulm, and/or Chemnitz Linuxtage.

In order to get SLS shipped with Free Software operating system distibutions,
maintainers of Mailman packages for e.g. GNU/Linux distributions (and the
Sabayon and Smallsister projects) will get asked (and offered help) to include
the patch.  The author will work with the Debian Mailman package maintainer to
try to get the patched Mailman shipped with Debian and Ubuntu, as discussed in
a private conversation with the maintainer, Tilburg, 2008-06-06.  The decision
on wether or not to include this patch is under control of the package
maintainer (not the patch author).

\end{document}

